Modal Abstractions for Virtualizing Memory Addresses

### 2024-06-28 examples.tex
+

As part of the full page table walk verification, we have verified the critical step for a small x86-64 kernel, which is the physical-to-virtual
conversion (often appearing as a macro \texttt{P2V} in C source code). In our small kernel, as in larger kernels, \texttt{P2V} is actually just addition,
but the correctness of this simple instruction is quite subtle, as shown in Figure \ref{fig:p2v}.
Beyond a frame ($P$), the precondition requires the current address space invariant, a check that the addition will not overflow, the physical address to convert
(in \lstinline|rbx|), knowledge that the current address space is valid, and knowledge that the physical address is in a range dedicated
to page table entries (as this physical-to-virtual conversion is only guaranteed for page table memory). 
The latter uses an additional ghost set added to the per-address-space
invariant per Figure \ref{fig:peraspaceinvariant_with_p2v_extension}.\todo{should we also discuss the following before while introducing the invariant?} The extended invariant owns, for each memory location dedicated to page tables, some fraction of
that physical memory, and also guarantees the offset is the correct calculation.
Subtly, the operation ensures the relevant table entry is readable, but the exact portion of ownership returned must be determined by inspecting the valid bit
of the value in memory --- so full ownership is returned only for unused entries.
This is a simple piece of code whose functionality is critical and whose correctness is highly non-trivial. No prior work engages with this problem.
\subsubsection{Traversing Page Tables}

\begin{figure}\footnotesize
\begin{lstlisting}[mathescape]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ulcorner \textsf{OFF} + \paddr > 0 \urcorner \ast \texttt{rbx}\mapsto_{\textsf{r}} \paddr \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}$
add rbx, OFF ;; shift physical address by identity mapping constant OFF
$\specline{\textsf{P} \ast \exists \qfraczero; \qfrac,\textsf{ w, val},  \texttt{rbx}\mapsto_{\textsf{r}} \paddr + \textsf{OFF} \ast \sumapaces\rtv\delta \ast \sumpv\paddr\qfrac\true \ast \vaddr \mapsto_{\texttt{vpte}} \{\textsf{q}, \textsf{qfrac}\} \paddr \;(\texttt{val})}$
$\specline{\ulcorner \vaddr = \paddr + \textsf{OFF} \urcorner \ast \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \ast \sumwalkabs\vaddr\qfrac\paddr \ast (R \wand \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)) }_{\rtv}$
\end{lstlisting}
\vspace{-1em}
\caption{Traversing page-tables, and allocating entries as needed while mapping-a-page in Figure \ref{fig:mappingcode}.}
\label{fig:walkpgdir}
\end{figure}

% \citet{kolanski08vstte,kolanski09tphols} verified a single code block with their logic which was roughly Figure \ref{fig:mapping_code} for a 2-level ARM
% page table, but several critical complexities our work deals with were not addressed.
% First, beyond the limitations discussed in Section \ref{sec:overly-restrictive}, Kolanski and Klein assumed that virtual addresses
% for page tables at each level were given as parameters rather than verifying any conversion from physical addresses to virtual addresses (or even axiomatizing their lookup).
% In contrast, our verification articulates the address space invariant from which the physical-to-virtual translation can be implemented.
% Second, our proof deals with the construction of a valid virtual points-to \emph{to the PTE to update} in mapping, which Kolanski and Klein also
% assumed was given.
% \todo{some of this is really an argument for our verification being more thorough, rather than being about our logic}

>>>>

As part of the full page table walk verification, we have verified the critical step for a small x86-64 kernel, which is the physical-to-virtual
conversion (often appearing as a macro \texttt{P2V} in C source code). In our small kernel, as in larger kernels, \texttt{P2V} is actually just addition,
but the correctness of this simple instruction is quite subtle, as shown in Figure \ref{fig:p2v}.
Beyond a frame ($P$), the precondition requires the current address space invariant, a check that the addition will not overflow, the physical address to convert
(in \lstinline|rbx|), knowledge that the current address space is valid, and knowledge that the physical address is in a range dedicated
to page table entries (as this physical-to-virtual conversion is only guaranteed for page table memory). 
The latter uses an additional ghost set added to the per-address-space
invariant per Figure \ref{fig:peraspaceinvariant_with_p2v_extension}. The extended invariant owns, for each memory location dedicated to page tables, some fraction of
that physical memory, and also guarantees the offset is the correct calculation.
Subtly, the operation ensures the relevant table entry is readable, but the exact portion of ownership returned must be determined by inspecting the valid bit
of the value in memory --- so full ownership is returned only for unused entries.
This is a simple piece of code whose functionality is critical and whose correctness is highly non-trivial. No prior work engages with this problem.
\subsubsection{Traversing Page Tables}
As we explain in Section \ref{}, a traversal over a page-table tree can be done through the indices into the tables at each level followed by obtaining the address of the next table. In Figure \ref{fig:getnexttable}, we see how to obtain a 
\begin{figure}\footnotesize
 \begin{lstlisting}[mathescape]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \ulcorner \textsf{OFF} + \paddr > 0 \urcorner \ast \texttt{rbx}\mapsto_{\textsf{r}} \paddr \ast \sumapaces\rtv\delta  \ast \sumpv\paddr\qfrac\true  }_{\rtv}$   
push   rbp
mov    rsp,rbp
sub    0x20,rsp
mov    rdi,-0x8[rbp]
;;  pte_t *next;

mov    -0x8[rbp],rdi
mov    [rdi],rdi
and    0x1,rdi
mov    rdi,rax

;;  if (!entry->present) {
cmp    0x0,rax
jne    161 <pte_get_next_table_succ+0xa1>

lea    -0x10[rbp],rdi

;;pte_initialize(&next);
callq  70 <pte_initialize>
mov    -0x8[rbp],rdi
mov    [rdi],rax

;; entry->writable = 1
;;entry->user_acc = 1;
;;entry->pfn = PTE_ADDR_TO_PHYS_PFN((uintptr_t) next);
;;entry->present = 1;
and    0xfffffffffffffffd,rax
or     0x2,rax
mov    rax,[rdi]
...
;;}

;;uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr);
mov    -0x8[rbp],rax
mov    [rax],rax
shr    0xc,rax
movabs 0xffffffffff,rcx
and    rcx,rax
shl    0xc,rax
mov    rax,-0x18[rbp]

;;uintptr_t next_virt_addr = (uintptr_t) P2V(next_phys_addr);
mov    -0x18[rbp],rax
movabs 0x0,rcx	187: R_X86_64_64	KERN_BASE
add    rcx,rax

;;next = (pte_t *) next_virt_addr;
mov    rax,-0x20[rbp]
mov    -0x20[rbp],rax
mov    rax,-0x10[rbp]

;;return next
mov    -0x10[rbp],rax
add    0x20,rsp
pop    rbp
ret
\end{lstlisting}
\vspace{-1em}
\caption{ in Figure \ref{fig:walkpgdir}.}
\label{fig:getnexttable}
\end{figure}

\begin{figure}\footnotesize
  \begin{lstlisting}[mathescape]
;;pte_t *walkpgdir_succ(pte_t *pml4, const void *va) {
push   rbp
mov    rsp,rbp
sub    0x40,rsp
mov    rdi,-0x8[rbp]
mov    rsi,-0x10[rbp]

;;pte_t *pml4_entry = &pml4[PML4EX(va)];
mov    -0x8[rbp],[rsi]
mov    -0x10[rbp],rdi
shr    0x27,rdi
and    0x1ff,rdi
shl    0x3,rdi
add    rdi,rsi
mov    rsi,-0x18[rbp]

;;pte_t *pdp = pte_get_next_table_succ(pml4_entry);
mov    -0x18[rbp],rdi
callq  <pte_get_next_table_succ>
mov    rax,-0x20[rbp]

;;pte_t *pdp_entry = &pdp[PDPEX(va)];
mov    -0x20[rbp],rax
mov    -0x10[rbp],rsi
shr    0x1e,rsi
and    0x1ff,rsi
shl    0x3,rsi
add    rsi,rax
mov    rax,-0x28[rbp]

;;pte_t *pd = pte_get_next_table_succ(pdp_entry);
mov    -0x28[rbp],rdi
callq  <pte_get_next_table_succ>
mov    rax,-0x30[rbp]

;;pte_t *pd_entry = &pd[PDEX(va)];
mov    -0x30[rbp],rax
mov    -0x10[rbp],rsi
shr    0x15,rsi
and    0x1ff,rsi
shl    0x3,rsi
add    rsi,rax
mov    rax,-0x38[rbp]

;;pte_t *pt = pte_get_next_table_succ(pd_entry);
mov    -0x38[rbp],rdi
callq  <pte_get_next_table_succ>
mov    rax,-0x40[rbp]

;;return &pt[PTEX(va)];
mov    -0x40[rbp],rax
mov    -0x10[rbp],rsi
shr    0xc,rsi
and    0x1ff,rsi
shl    0x3,rsi
add    rsi,rax
add    0x40,rsp
pop    rbp
retq   
\end{lstlisting}
\vspace{-1em}
\caption{Traversing page-tables, and allocating entries as needed while mapping-a-page in Figure \ref{fig:mappingcode}.}
\label{fig:walkpgdir}
\end{figure}
% \citet{kolanski08vstte,kolanski09tphols} verified a single code block with their logic which was roughly Figure \ref{fig:mapping_code} for a 2-level ARM
% page table, but several critical complexities our work deals with were not addressed.
% First, beyond the limitations discussed in Section \ref{sec:overly-restrictive}, Kolanski and Klein assumed that virtual addresses
% for page tables at each level were given as parameters rather than verifying any conversion from physical addresses to virtual addresses (or even axiomatizing their lookup).
% In contrast, our verification articulates the address space invariant from which the physical-to-virtual translation can be implemented.
% Second, our proof deals with the construction of a valid virtual points-to \emph{to the PTE to update} in mapping, which Kolanski and Klein also
% assumed was given.
% \todo{some of this is really an argument for our verification being more thorough, rather than being about our logic}
